Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-23 paper/geometry-twocolumn.tex
Update the geometry section

We motivate our investigation of algebraically indexed types and their
relational interpretations by developing a novel type system for
programs that manipulate two-dimensional geometric data. Geometry is
rich with operations that are invariant under transformation: affine
operations are invariant under change of origin, vector space
operations are invariant under change of basis, and dot product is
invariant under orthogonal changes of basis. On the other hand, some
geometric operations are interestingly variant under transformation.
For example, cross products vary with scalings of the plane. We
incorporate (in)variance information about geometric primitives
into type systems via algebraically indexed types.

@@

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
we assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The (simplified) index-erasure and relation interpretations
of $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{\rho'} \sepbar \rho' \in
    \extends{\rho}{t\mathord:\SynTransl{2}} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
depends on which (possibly non-compositional) interpretation of
relational environments we use. For now we can think of environments
as mappings from free variables of sort $\SynTransl{2}$ to the group
$\Transl{2}$, and of extension as adding extra components to
mappings. The index-erasure and relational interpretations are given
formally in Sections~\ref{sec:index-erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\ref{sec:relational-semantics}. %, respectively.

@@

Invariance under change of origin is the key feature of affine
geometry, whose central operation is the affine combination of points:
$\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
\lambda_2 = 1$.  Geometrically, this can be interpreted as
interpolation between the points represented by $\vec{v_1}$ and
$\vec{v_2}$.  We add affine combination of points to our calculus as
follows: % typing and intended denotation:for
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord:
    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
    r\vec{v_2} 
\end{array}
\end{displaymath}
It can be verified by hand that %the intended denotation
$\sem{\mathrm{affComb}}$ is invariant under all changes of origin, as
dictated by its type.
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

@@

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  quantifier-free inhabitants\footnote{We define quantifier-free
    programs in the statement of
    \thmref{thm:abstraction}.}. Intuitively, this is because we cannot
  remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
  vector operations. We formalise this indefinability result in
  \autoref{sec:types-indexed-abelian-groups-indef} using a more
  sophisticated relational interpretation of free translation
  variables than the one sketched above.
\end{example}

@@

Although vector addition, negation, and scaling are not invariant
under change of origin, they are invariant under change of basis. As
with origin invariance, we can express basis invariance as
preservation of relations indexed by changes of basis. Change of basis
is achieved by applying an invertible linear map, and the collection of
all such maps on $\mathbb{R}^2$ forms the \emph{General Linear} group
$\GL{2}$ which we represent in our language by a new indexing sort
$\SynGL{2}$ with (non-abelian) group structure that we will write
multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and (simplified)
relational semantics of $\tyPrim{vec}{B,t}$ are given by:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

\paragraph{Affine Geometry} An \emph{affine transformation} 
an invertible linear map together with a translation. We can assign
types to all the primitive affine and vector space operations
indicating how they they behave with respect to affine
transformations:
\begin{eqnarray*}
  \mathrm{affComb} & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\\
    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
  \end{array}
  \\
  (+) & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
    \tyPrim{vec}{B,t_1 + t_2}
  \end{array}
  \\
  \mathrm{negate} & : & \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  0 & : & \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
  (*) & : & \forall B \mathord: \SynGL{2}.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
\end{eqnarray*}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the \emph{dot product}, or \emph{inner product},
operation of two vectors. The dot product is defined 
%on the coordinate representation 
by
%\begin{displaymath}
$  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2$.
%\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To %incorporate the operation of dot product into our calculus, we
must assign it a type we note that, although dot product is not
invariant under $\GL{2}$ or $\Transl{2}$, it
% that we have considered so far. However,e  the dot product 
is invariant under the subgroup $\Orth{2}$ of $\GL{2}$ of
\emph{orthogonal} linear transformations, i.e., the subgroup of
invertible linear maps whose matrix representations' transposes are
equal to their inverses. We thus introduce a new sort $\SynOrth{2}$ of
orthogonal transformations, and overload the multiplicative group
operations for inhabitants of $\SynOrth{2}$. Further assuming a
function $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
\SynGL{2}$ we assign dot product this type:
%a type to the dot product, using quantification over $\SynOrth{2}$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

@@

The cross product of two vectors is defined on coordinate
representations as $(x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1$.
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors.  Under change of
basis by an invertible linear transformation $B$, the cross product of
two vectors varies with the determinant of $B$. This corresponds to
scaling the plane by the change of basis transformation, so we augment
our calculus with a new sort $\SynGL{1}$ of scale factors
(i.e.,~$1$-dimensional invertible linear maps). Semantically,
$\SynGL{1}$ ranges over the non-zero real numbers and forms an abelian
group which we write multiplicatively. We also add two new operations:
determinant, $\det B$, which takes an inhabitant of $\SynGL{2}$ to its
determinant in $\SynGL{1}$, and absolute value, $\abs{e}$, which takes
scaling factors to scaling factors. We also refine the type
$\tyPrimNm{real}$ of real numbers so that it is indexed by the sort
$\SynGL{1}$: $\tyPrim{real}{e}$. The old type $\tyPrimNm{real}$ is
then just $\tyPrim{real}{1}$, and the full collection of operations on
real numbers indexed by scaling factors is shown in
\autoref{fig:real-ops}. We can thus assign cross product the type:
%With the additional sort $\SynGL{1}$ and the refined type of real
%numbers, we can assign a type to the cross product:
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
\end{displaymath}
Since the absolute value of the determinant of an orthogonal
transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$ to
hold.

@@

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. We thus require
$(\iota_1 s)B = B(\iota_1 s)$ to hold.
The scaling maps $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
are precisely the elements of $\GL{2}$ that commute with all others;
these form the \emph{centre} of $\GL{2}$. If we keep track of
scalings, then we can assign the following more precise types to
scalar multiplcation and dot product:
\begin{displaymath}
\begin{array}{@{}l@{}l}
  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{\iota_1(s)B,0}\\
%\end{displaymath}
%and dot product:
%\begin{displaymath}
%  \begin{array}{l@{}l}
    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
  \end{array}
\end{displaymath}

\begin{example}
  With the operations in \autoref{fig:real-ops}, it is not possible to
  write a quantifier-free program with the following type that is not
  constantly zero:
  \begin{displaymath}
    \forall s \mathord: \SynGL{2}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
  \end{displaymath}
  This was shown by Kennedy for his units of measure system
  \cite{kennedy97relational}.  In particular, it is not possible to
  write a square root function with the above type.  The
  indefinability of square root is similar to the uninhabitation of the
  type in \exref{ex:uninhabited-type}.

  In \autoref{sec:types-indexed-abelian-groups-indef} we revisit
  Kennedy's result and show that even if we add square root as a
  primitive operation---with the type above---then it is still not
  possible to construct the cube root function. The indefinability of
  cube root is related to the impossibility of trisecting an arbitrary
  angle by ruler and compass constructions.
%  is of interest due to its relevance to the classical problem of
%  trisecting an angle by ruler and compass constructions.
\end{example}

>>>>

We motivate our investigation of algebraically indexed types and their
relational interpretations by developing a novel type system for
programs that manipulate two-dimensional geometric data. Geometry is
rich with operations that are invariant under transformation: affine
operations are invariant under change of origin
(\autoref{sec:affine-vector-ops}), vector space operations are
invariant under change of basis, and dot product is invariant under
orthogonal changes of basis
(\autoref{sec:motivation-generalising}). On the other hand, some
geometric operations are interestingly variant under transformation.
For example, cross products vary with scalings of the plane
(\autoref{sec:scale-invariance}). We incorporate (in)variance
information about geometric primitives into type systems via
algebraically indexed types.

@@

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$, we
assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The index-erasure and relational interpretations of
$\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{(\rho,\vec{t})} \sepbar \vec{t} \in \Transl{2} \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpretations are given
formally in Sections~\ref{sec:erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\ref{sec:relational-semantics}. %, respectively.

@@

Invariance under change of origin is the key feature of affine
geometry, whose central operation is the affine combination of points:
$\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
\lambda_2 = 1$.  Geometrically, this can be interpreted as describing
all the points on the unique line through the points represented by
$\vec{v_1}$ and $\vec{v_2}$ (assuming $\vec{v_1} \not= \vec{v_2}$).
We add affine combination of points to our calculus as
follows: % typing and intended denotation:for
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord:
    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
    r\vec{v_2} 
\end{array}
\end{displaymath}
It can be verified by hand that %the intended denotation
$\sem{\mathrm{affComb}}$ is invariant under all changes of origin, as
dictated by its type.
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

@@

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  inhabitants. Intuitively, this is because we cannot remove the extra
  $t$ in $\tyPrim{vec}{t+t}$ using the vector operations. We formalise
  this non-definability result in \autoref{sec:general-nondef} using a
  specialised relational interpretation.
\end{example}

@@

Although vector addition, negation, and scaling are not invariant
under change of origin, they are invariant under change of basis. As
with origin invariance, we can express basis invariance as
preservation of relations indexed by changes of basis. Change of basis
is achieved by applying an invertible linear map, and the collection
of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
group $\GL{2}$ which we represent in our language by a new indexing
sort $\SynGL{2}$ with (non-abelian) group structure that we will write
multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and relational
semantics of $\tyPrim{vec}{B,t}$ are given by:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

\paragraph{Affine Geometry} An \emph{affine transformation} is
an invertible linear map together with a translation. We can assign
types to all the primitive affine and vector space operations
indicating how they they behave with respect to affine
transformations:
\begin{eqnarray*}
  \mathrm{affComb} & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\\
    \hspace{0.2cm} \tyPrim{vec}{B,t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,t}
  \end{array}
  \\
  (+) & : &
  \begin{array}[t]{@{}l}
    \forall B \mathord: \SynGL{2}, t_1,t_2 \mathord: \SynTransl{2}. \\
    \hspace{0.2cm}\tyPrim{vec}{B,t_1} \to \tyPrim{vec}{B,t_2} \to
    \tyPrim{vec}{B,t_1 + t_2}
  \end{array}
  \\
  \mathrm{negate} & : & \forall B \mathord: \SynGL{2}, t \mathord: \SynTransl{2}.\ \tyPrim{vec}{B,t} \to \tyPrim{vec}{B,-t} \\
  0 & : & \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \\
  (*) & : & \forall B \mathord: \SynGL{2}.\ \tyPrimNm{real} \to \tyPrim{vec}{B,0} \to \tyPrim{vec}{B,0}
\end{eqnarray*}

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the \emph{dot product}, or \emph{inner product},
operation of two vectors. The dot product is defined 
%on the coordinate representation 
by
%\begin{displaymath}
$  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2$.
%\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To %incorporate the operation of dot product into our calculus, we
assign it a type we note that, although dot product is not
invariant under $\GL{2}$ or $\Transl{2}$, it
% that we have considered so far. However,e  the dot product 
is invariant under the subgroup $\Orth{2}$ of $\GL{2}$ of
\emph{orthogonal} linear transformations, i.e., the subgroup of
invertible linear maps whose matrix representations' transposes are
equal to their inverses. We thus introduce a new sort $\SynOrth{2}$ of
orthogonal transformations, and overload the multiplicative group
operations for inhabitants of $\SynOrth{2}$. Further assuming an
injection $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
\SynGL{2}$ we assign dot product this type:
%a type to the dot product, using quantification over $\SynOrth{2}$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

@@

The cross product of two vectors is defined on coordinate
representations as $(x_1,y_1) \times (x_2,y_2) = x_1y_2 - x_2y_1$.
Geometrically, the cross product is the signed area of the
parallelogram described by the pair of input vectors.  Under change of
basis by an invertible linear transformation $B$, the cross product of
two vectors varies with the determinant of $B$. This corresponds to
scaling the plane by the change of basis transformation, so we augment
our calculus with a new sort $\SynGL{1}$ of scale factors
(i.e.,~$1$-dimensional invertible linear maps). Semantically,
$\SynGL{1}$ ranges over the non-zero real numbers and forms an abelian
group which we write multiplicatively. We also add two new operations:
determinant, $\det B$, which takes an inhabitant of $\SynGL{2}$ to its
determinant in $\SynGL{1}$, and absolute value, $\abs{e}$, which takes
scaling factors to scaling factors. We also refine the type
$\tyPrimNm{real}$ of real numbers so that it is indexed by the sort
$\SynGL{1}$: $\tyPrim{real}{e}$. The old type $\tyPrimNm{real}$ is
then just $\tyPrim{real}{1}$, and the full collection of operations on
real numbers indexed by scaling factors is shown in
\autoref{fig:real-ops}. We can thus assign cross product the type:
%With the additional sort $\SynGL{1}$ and the refined type of real
%numbers, we can assign a type to the cross product:
\begin{displaymath}
  (\times) : \forall B \mathord: \SynGL{2}.\ \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{B,0} \to \tyPrim{real}{\det B} 
\end{displaymath}
Since the absolute value of the determinant of an orthogonal
transformation is always $1$, we assume $\abs{\det (\iota_O O)} = 1$
to hold for any $O \in \SynOrth{2}$.

@@

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. We thus require
$(\iota_1 s)B = B(\iota_1 s)$ to hold.
The scaling maps $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$
are precisely the elements of $\GL{2}$ that commute with all others;
these form the \emph{centre} of $\GL{2}$. If we keep track of
scalings, then we can assign the following more precise types to
scalar multiplication and dot product:
\begin{displaymath}
\begin{array}{@{}l@{}l}
  (*) : \forall s \mathord: \SynGL{1}, B \mathord:
  \SynGL{2}.\ & \tyPrim{real}{s} \to \tyPrim{vec}{B,0} \to
  \tyPrim{vec}{\iota_1(s)B,0}\\
%\end{displaymath}
%and dot product:
%\begin{displaymath}
%  \begin{array}{l@{}l}
    (\cdot) : \forall s \mathord: \SynGL{1}, O \mathord: \SynOrth{2}.\ & \tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to\\
    &\tyPrim{vec}{\iota_1(s)\iota_O(O),0} \to \tyPrim{real}{s^2}
  \end{array}
\end{displaymath}

\begin{example}
  With the operations in \autoref{fig:real-ops}, it is not possible to
  write a term with the following type that is not constantly zero:
  \begin{displaymath}
    \forall s \mathord: \SynGL{2}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}
  \end{displaymath}
  This was shown by Kennedy for his units of measure system
  \cite{kennedy97relational}.  In particular, it is not possible to
  write a square root function with the above type.  The
  non-definability of square root is similar to the uninhabitation of the
  type in \exref{ex:uninhabited-type}.

  In \autoref{sec:types-indexed-abelian-groups-indef} we revisit
  Kennedy's result and show that even if we add square root as a
  primitive operation---with the type above---then it is still not
  possible to construct the cube root function. The non-definability of
  cube root is related to the impossibility of trisecting an arbitrary
  angle by ruler and compass constructions.
%  is of interest due to its relevance to the classical problem of
%  trisecting an angle by ruler and compass constructions.
\end{example}
